### 2019-11-01 latex/vnhcs.tex
Changed the bibliography style This new version changes the bibliography type to use a different package, and we now compile with lualatex. It helps with error and warning messages, as it actually shows where the message is coming from for proper errors (as opposed to always doing it in main)

%/===============/%
\section{Virtual Nonholonomic Constraints}

@@

%/----- Lagrange vs Hamilton -----/%
\subsection{A Hamiltonian Approach to Virtual Nonholonomic Constraints}
Take an underactuated mechanical system 
\begin{equation}\label{eqn:eom_lagrangian}
M(q)\ddot{q} + C(q,\dot{q})\dot{q} + \nabla V(q) = B(q)\tau
\end{equation}
with Lagrangian
\begin{equation}\label{eqn:lagrangian}
\mathcal{L}(q,\dot{q}) = \frac{1}{2} \dot{q}^T M(q) \dot{q} - V(q)
\end{equation}
where \(q = (q_1,\ldots,q_n) \in \mathcal{Q}\) is the configuration vector, the inertia matrix \(M(q) = M^T(q)\) is positive definite for all \(q \in \mathcal{Q}\), \(V(q)\) is the potential of the system, \(B : \mathcal{Q} \rightarrow \mathbb{R}^{n \times m}\) is a \(C^1\) full rank control matrix, and \(\tau \in \mathbb{R}^{m}\) is the control input.

One can compute the Hamiltonian of this system by performing a Legendre transform on the Lagrangian. Define the conjugate of momenta for \(q\) by 
\[
p = \frac{\delta \mathcal{L}}{\delta \dot{q}} \in \mathbb{R}^n
\]
Then, the Legendre transform is performed by taking
\[
\mathcal{H}(q,p) = p^T \dot{q} - \mathcal{L}(q,\dot{q})
\]
It is easy to verify that the Hamiltonian for system (\ref{eqn:lagrangian}) is given by the total mechanical energy (\ref{eqn:hamiltonian}).
\begin{equation}\label{eqn:hamiltonian}
\mathcal{H}(q,p) = E(q,p) = \frac{1}{2} p^T M^{-1}(q) p + V(q)
\end{equation}

The equations of motion for the system in Hamiltonian coordinates is given by
\begin{align}\label{eqn:hamiltionian_eom}
\begin{split}
\dot{q} &= \frac{\delta \mathcal{H}}{\delta p} = M^{-1}(q) p \\
\dot{p} &= -\frac{\delta \mathcal{H}}{\delta q} + B \tau
\end{split}
\end{align}

Suppose the mechanical system has degree of underactuation one, so that coordinates of the system can be split into an unactuated component \(q_u \in [\mathbb{R}]_T, \, T \in \mathbb{R}_{>0}\) which is not influence by control, along with an actuated component \(q_a\); that is, suppose \(B\) is of the form \(B(q) = [0_m, B_1^T(q), \ldots, B_n^T(q)]^T, \, B_i^T(q) \in \mathbb{R}^{n-1}\) and \(\tau \in \mathbb{R}^{n-1}\). In this case, \(q = (q_u, q_a)^T\) and the equations of motion become
\begin{align}\label{eqn:unactuated_actuated_eom}
\begin{split}
\dot{q_u} &= e_1^T M^{-1}(q) p \\
\dot{p_u} &= -p^T\frac{\delta M}{\delta q_u} p - \delta_{q_u}V(q) \\
\dot{q_a} &= 
\begin{bmatrix}
0 \cdots 0 \\
I_{n-1} \\
\end{bmatrix} M^{-1}(q) p \\
\dot{p_a} &= -p^T\frac{\delta M}{\delta q_a} p - \nabla{q_a}V(q) + 
\begin{bmatrix}
B_1(q) \\
\vdots \\
B_n(q)
\end{bmatrix} \tau
\end{split}
\end{align}

@@

Now we solve for \(\tau\) by finding \(\ddot{e}\).
\begin{align*}
    e &= h(q,p) = qa - f(q_u,p_u)\\
    \Rightarrow \dot{e} &= \dot{q_a} - df_{q_u}\dot{q_u} -df_{p_u}\dot{p_u} \\
    &= [-df_{q_u} I_{n-1}]\dot{q} - df_{p_u} \dot{p_u} \\
    &= dh_q M^{-1}(q) p - df_{p_u}\left( -\frac{1}{2}p^T \frac{\delta M^{-1}(q)}{\delta q_u} p - \delta_{q_u}V(q) \right) \\
    &= dh_q M^{-1}(q) p + \frac{1}{2}df_{p_u} p^T \frac{\delta M^{-1}(q)}{\delta q_u} p + df_{p_u}\delta_{q_u}V(q)
\end{align*}
The control input \(\tau\) only appears in \(\dot{p_a}\) (see (\ref{eqn:unactuated_actuated_eom})). To simplify the analysis, terms in \(\ddot{e}\) which do not depend on \(\dot{p_a}\) explicitly are lumped together under the symbol \((*)\):
\begin{align*}
    \ddot{e} &= dh_q M^{-1}(q) \dot{p} + df_{p_u} p^T \frac{\delta M^{-1}(q)}{\delta q_u} \dot{p} + (*) \\
    &= (dh_q M^{-1}(q) + df_{p_u} p^T \frac{\delta M^{-1}(q)}{\delta q_u})B\tau + (*) \\
    &= (dh_q M^{-1}(q) + dh_{p_u} p^T \frac{\delta M^{-1}(q)}{\delta q_u})B\tau + (*)
\end{align*}

From the derivations above, one can solve for \(\tau\) iff the matrix on the left of \(\tau\) is full rank. Thus, for systems with degree of underactuation one we give the following definition.
\begin{defn}
A VNHC \(h(q,p) = 0\) of order \(n - 1\) is \textbf{regular} if \(dh_{p_a} = 0\), \(dh_{q_a} = (1 \ldots 1)^T\), and 
\[
\text{rank}\left\{ (dh_q M^{-1}(q) + dh_{p_u} p^T \frac{\delta M^{-1}(q)}{\delta q_u})B\right\} = n - 1
\]
everywhere on the constraint manifold \(\Gamma\). Equivalently, a VNHC \(h\) of order \(n - 1\) is regular if it satisfies Assumption \ref{assm:vnhc_is_on_qu_pu} and system (\ref{eqn:unactuated_actuated_eom}) with output \(e = h(q,p)\) is of relative degree \(\{2,2,\ldots,2\}\) everywhere on \(\Gamma\).
\end{defn}

In general, \(\dot{e}\) is a function of \(q_u\) and \(p = (p_u,p_a)^T\). Since the purpose of a regular VNHC is to fully parameterize \(\Gamma\) by \((q_u,p_u)\), it is essential that one can solve for \(p_a = p_a(q_u,p_u)\). Unfortunately this often cannot be done, since \(\dot{e}\) contains the quadratic term
\[
\frac{1}{2} df_{p_u} p^T \frac{\delta M^{-1}(q)}{\delta q_u} p 
\]
We can solve for \(p_a\) if this quadratic term does not exist.
\begin{assm}\label{assm:M_is_Mqa}
Assume \(\delta M(q) / \delta q_u = 0 \Leftrightarrow \delta M^{-1}(q) / \delta q_u = 0\)
\end{assm}

@@

Now we solve for \(p_a\) on the constraint manifold (when \(e = \dot{e} = 0\)):
\begin{align*}
    \dot{e} = dh_q M^{-1}(q)p + df_{p_u} \delta_{q_u}V(q) &= 0\\
    \Leftrightarrow dh_q M^{-1}(q)e_1 p_u + dh_q M^{-1}(q) \begin{bmatrix}
    0 & \cdots & 0 \\
    & I_{n-1} & \\
    \end{bmatrix} p_a &= -df_{p_u} \delta_{q_u}V(q) \\
\end{align*}
\begin{align*}
    \Leftrightarrow dh_q M^{-1}(q) \begin{bmatrix}
    0 & \cdots & 0 \\
    & I_{n-1} & \\
    \end{bmatrix} p_a = -\left(df_{p_u}\delta_{q_u}V(q) + dh_q M^{-1}(q)e_1 p_u\right)
\end{align*}
One can linearly solve for \(p_a\) if and only if the matrix in front of it is invertible.

@@

\begin{thm}\label{thm:equation_for_pa}
Suppose assumptions \ref{assm:vnhc_is_on_qu_pu} and \ref{assm:M_is_Mqa} hold. If a regular VNHC \(h(q,p) = q_a - f(q_u,p_u)\) is solvable, then the parameterization for \(p_a\) on the constraint manifold is given by
\begin{align*}
p_a &= -\left(dh_q M^{-1}(q) \begin{bmatrix}
    0 & \cdots & 0 \\
    & I_{n-1} & \\
    \end{bmatrix}\right)^{-1}\left( df_{p_u}\delta_{q_u}V(q) + dh_q M^{-1}(q)e_1 p_u\right) \\
    &=: g(q_u,p_u)
\end{align*}
and the constrained dynamics on \(\Gamma\) are given by
\begin{align*}
    \dot{q_u} &= e_1^T M^{-1}(q_a) \begin{bmatrix}
    p_u \\
    p_a
    \end{bmatrix}\mid_{q_a = f(q_u,p_u), p_a = g(q_u,p_u)} \\
    \dot{p_u} &= -\delta_{q_u} V(q_u,q_a) \mid_{q_a = f(q_u,p_u)}
\end{align*}
\end{thm}

Theorem \ref{thm:equation_for_pa} guarantees that, on \(\Gamma\), \(q_a\) is a parameterized completely by \((q_u,p_u)\). Hence, the zero-dynamics on \(\Gamma\) are always two-dimensional regardless of the original dimension \(n\).
%/===============/%

>>>>

%! TEX root = main.tex

%/========== Virtual Nonholonomic Constraints ==========/%
\section{Virtual Nonholonomic Constraints}

@@

%/----- Lagrange vs Hamilton -----/%
\subsection{A Hamiltonian Approach to Virtual Nonholonomic Constraints}
Take an underactuated mechanical system

\begin{equation}\label{eqn:eom_lagrangian}
M(q)\ddot{q} + C(q,\dot{q})\dot{q} + \nabla V(q) = B(q)\tau
\end{equation}
with Lagrangian
\begin{equation}\label{eqn:lagrangian}
\mathcal{L}(q,\dot{q}) = \frac{1}{2} \dot{q}^T M(q) \dot{q} - V(q)
\end{equation}
where \(q = (q_1,\ldots,q_n) \in \mathcal{Q}\) is the configuration vector, the inertia matrix \(M(q) = M^T(q)\) is positive definite for all \(q \in \mathcal{Q}\), \(V(q)\) is the potential of the system, \(B : \mathcal{Q} \rightarrow \mathbb{R}^{n \times m}\) is a \(C^1\) full rank control matrix, and \(\tau \in \mathbb{R}^{m}\) is the control input.

One can compute the Hamiltonian of this system by performing a Legendre transform on the Lagrangian. Define the conjugate of momenta for \(q\) by 
\[
p = \frac{\partial \mathcal{L}}{\partial \dot{q}} \in \mathbb{R}^n
\]
Then, the Legendre transform is performed by taking
\[
\mathcal{H}(q,p) = p^T \dot{q} - \mathcal{L}(q,\dot{q})
\]
It is easy to verify that the Hamiltonian for system (\ref{eqn:lagrangian}) is given by the total mechanical energy (\ref{eqn:hamiltonian}).
\begin{equation}\label{eqn:hamiltonian}
\mathcal{H}(q,p) = E(q,p) = \frac{1}{2} p^T M^{-1}(q) p + V(q)
\end{equation}

The equations of motion for the system in Hamiltonian coordinates is given by
\begin{align}\label{eqn:hamiltionian_eom}
\begin{split}
\dot{q} &= \frac{\partial \mathcal{H}}{\partial p} = M^{-1}(q) p \\
\dot{p} &= -\frac{\partial \mathcal{H}}{\partial q} + B \tau
\end{split}
\end{align}

Suppose the mechanical system has degree of underactuation one, so that coordinates of the system can be split into an unactuated component \(q_u \in [\mathbb{R}]_T, \, T \in \mathbb{R}_{>0}\) which is not influence by control, along with an actuated component \(q_a\); that is, suppose \(B\) is of the form \(B(q) = [0_m, B_1^T(q), \ldots, B_n^T(q)]^T, \, B_i^T(q) \in \mathbb{R}^{n-1}\) and \(\tau \in \mathbb{R}^{n-1}\). In this case, \(q = (q_u, q_a)^T\) and the equations of motion become
\begin{align}\label{eqn:unactuated_actuated_eom}
\begin{split}
\dot{q_u} &= e_1^T M^{-1}(q) p \\
\dot{p_u} &= -p^T\frac{\partial M}{\partial q_u} p - \partial_{q_u}V(q) \\
\dot{q_a} &= 
\begin{bmatrix}
0 \cdots 0 \\
I_{n-1} \\
\end{bmatrix} M^{-1}(q) p \\
\dot{p_a} &= -p^T\frac{\partial M}{\partial q_a} p - \nabla{q_a}V(q) + 
\begin{bmatrix}
B_1(q) \\
\vdots \\
B_n(q)
\end{bmatrix} \tau
\end{split}
\end{align}

@@

Now we solve for \(\tau\) by finding \(\ddot{e}\).
\begin{align*}
    e &= h(q,p) = qa - f(q_u,p_u)\\
    \Rightarrow \dot{e} &= \dot{q_a} - df_{q_u}\dot{q_u} -df_{p_u}\dot{p_u} \\
    &= [-df_{q_u} I_{n-1}]\dot{q} - df_{p_u} \dot{p_u} \\
    &= dh_q M^{-1}(q) p - df_{p_u}\left( -\frac{1}{2}p^T \frac{\partial M^{-1}(q)}{\partial q_u} p - \partial_{q_u}V(q) \right) \\
    &= dh_q M^{-1}(q) p + \frac{1}{2}df_{p_u} p^T \frac{\partial M^{-1}(q)}{\partial q_u} p + df_{p_u}\partial_{q_u}V(q)
\end{align*}
The control input \(\tau\) only appears in \(\dot{p_a}\) (see (\ref{eqn:unactuated_actuated_eom})). To simplify the analysis, terms in \(\ddot{e}\) which do not depend on \(\dot{p_a}\) explicitly are lumped together under the symbol \((*)\):
\begin{align*}
    \ddot{e} &= dh_q M^{-1}(q) \dot{p} + df_{p_u} p^T \frac{\partial M^{-1}(q)}{\partial q_u} \dot{p} + (*) \\
    &= (dh_q M^{-1}(q) + df_{p_u} p^T \frac{\partial M^{-1}(q)}{\partial q_u})B\tau + (*) \\
    &= (dh_q M^{-1}(q) + dh_{p_u} p^T \frac{\partial M^{-1}(q)}{\partial q_u})B\tau + (*)
\end{align*}

From the derivations above, one can solve for \(\tau\) iff the matrix on the left of \(\tau\) is full rank. Thus, for systems with degree of underactuation one we give the following definition.
\begin{defn}
A VNHC \(h(q,p) = 0\) of order \(n - 1\) is \textbf{regular} if \(dh_{p_a} = 0\), \(dh_{q_a} = (1 \ldots 1)^T\), and 
\[
\text{rank}\left\{ (dh_q M^{-1}(q) + dh_{p_u} p^T \frac{\partial M^{-1}(q)}{\partial q_u})B\right\} = n - 1
\]
everywhere on the constraint manifold \(\Gamma\). Equivalently, a VNHC \(h\) of order \(n - 1\) is regular if it satisfies Assumption \ref{assm:vnhc_is_on_qu_pu} and system (\ref{eqn:unactuated_actuated_eom}) with output \(e = h(q,p)\) is of relative degree \(\{2,2,\ldots,2\}\) everywhere on \(\Gamma\).
\end{defn}

In general, \(\dot{e}\) is a function of \(q_u\) and \(p = (p_u,p_a)^T\). Since the purpose of a regular VNHC is to fully parameterize \(\Gamma\) by \((q_u,p_u)\), it is essential that one can solve for \(p_a = p_a(q_u,p_u)\). Unfortunately this often cannot be done, since \(\dot{e}\) contains the quadratic term
\[
\frac{1}{2} df_{p_u} p^T \frac{\partial M^{-1}(q)}{\partial q_u} p 
\]
We can solve for \(p_a\) if this quadratic term does not exist.
\begin{assm}\label{assm:M_is_Mqa}
Assume \(\partial M(q) / \partial q_u = 0 \Leftrightarrow \partial M^{-1}(q) / \partial q_u = 0\)
\end{assm}

@@

Now we solve for \(p_a\) on the constraint manifold (when \(e = \dot{e} = 0\)):
\begin{align*}
    \dot{e} = dh_q M^{-1}(q)p + df_{p_u} \partial_{q_u}V(q) &= 0\\
    \Leftrightarrow dh_q M^{-1}(q)e_1 p_u + dh_q M^{-1}(q) \begin{bmatrix}
    0 & \cdots & 0 \\
    & I_{n-1} & \\
    \end{bmatrix} p_a &= -df_{p_u} \partial_{q_u}V(q) \\
\end{align*}
\begin{align*}
    \Leftrightarrow dh_q M^{-1}(q) \begin{bmatrix}
    0 & \cdots & 0 \\
    & I_{n-1} & \\
    \end{bmatrix} p_a = -\left(df_{p_u}\partial_{q_u}V(q) + dh_q M^{-1}(q)e_1 p_u\right)
\end{align*}
One can linearly solve for \(p_a\) if and only if the matrix in front of it is invertible.

@@

\begin{thm}\label{thm:equation_for_pa}
Suppose assumptions \ref{assm:vnhc_is_on_qu_pu} and \ref{assm:M_is_Mqa} hold. If a regular VNHC \(h(q,p) = q_a - f(q_u,p_u)\) is solvable, then the parameterization for \(p_a\) on the constraint manifold is given by
\begin{align*}
p_a &= -\left(dh_q M^{-1}(q) \begin{bmatrix}
    0 & \cdots & 0 \\
    & I_{n-1} & \\
    \end{bmatrix}\right)^{-1}\left( df_{p_u}\partial_{q_u}V(q) + dh_q M^{-1}(q)e_1 p_u\right) \\
    &=: g(q_u,p_u)
\end{align*}
and the constrained dynamics on \(\Gamma\) are given by
\begin{align*}
    \dot{q_u} &= e_1^T M^{-1}(q_a) \begin{bmatrix}
    p_u \\
    p_a
    \end{bmatrix}\mid_{q_a = f(q_u,p_u), p_a = g(q_u,p_u)} \\
    \dot{p_u} &= -\partial_{q_u} V(q_u,q_a) \mid_{q_a = f(q_u,p_u)}
\end{align*}
\end{thm}

Theorem \ref{thm:equation_for_pa} guarantees that, on \(\Gamma\), \(q_a\) is a parameterized completely by \((q_u,p_u)\). Hence, the zero-dynamics on \(\Gamma\) are always two-dimensional regardless of the original dimension \(n\).
%/========== /Virtual Nonholonomic Constraints ==========/%
